---
title: "R Notebook"
output:
  html_document: default
  html_notebook: default
---

**Principles of Data Visualization and Introduction to ggplot2**

I have provided you with data about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine. lets read this in:

```{r}

inc <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module1/Data/inc5000_data.csv", header= TRUE)

```

And lets preview this data:

```{r}

head(inc)

```
```{r}

summary(inc)

```

Think a bit on what these summaries mean. Use the space below to add some more relevant non-visual exploratory information you think helps you understand this data:

```{r echo = FALSE, message = FALSE, warning = FALSE}

# Load libraries to aid with intrepretation and visualization
library(ggplot2)
library(tidyr)
library(dplyr)
library(forcats)
library(scales)

```

```{r warning = FALSE}

# Rank US cities by number of companies in dataset
inc %>% 
  select(City) %>% 
  group_by(City) %>% 
  tally() %>% 
  arrange(desc(n)) %>% 
  head(10)

```

Interestingly, a number of [top ten population cities](https://en.wikipedia.org/wiki/List_of_United_States_cities_by_population) don't rank in top 10 in list.

- Los Angeles (#2 population ranking)
- Phoenix (#5)
- Philadelphia (#6)
- San Antonio (#7)
- San Jose (#10)

***

## Question 1

Create a graph that shows the distribution of companies in the dataset by State (ie how many are in each state). There are a lot of States, so consider which axis you should use. This visualization is ultimately going to be consumed on a 'portrait' oriented screen (ie taller than wide), which should further guide your layout choices.

```{r fig.width = 9/2, fig.height = 8}

# We chart states with horizontal bars representing the number of companies for each.  Top states are highlighted for ease of reference.

# Define custom monochromatic color gradient (hue, really) based on rank grouping to highlight states with most companies in the dataset - the top 3 are dark gray, the next seven are gray, and the remainder are light gray.
state_colors <- c('Top3' = 'grey30', '7-10' = 'grey55', '11-50' = 'grey84')

# Configure data for ggplot
inc %>% 
  select(Name, State) %>% 
  group_by(State) %>% 
  tally() %>% 
  
  # Create variables to control bar hue by state rank groupings - references: https://stackoverflow.com/questions/26106408/create-a-ranking-variable-with-dplyr and https://community.rstudio.com/t/setting-colours-in-ggplot-conditional-on-value/8328
  mutate(Ranking = dense_rank(desc(n)), 
         RankGroup = cut(Ranking,
                          breaks = c(0, 3, 10, Inf),
                          labels = c('Top3', '7-10', '11-50'),
                          right = TRUE)) %>% 

  # Call ggplot
  ggplot(aes(x = fct_rev(State),  # Order states reverse-alphabetically before flipping coordinate axes
             y = n,
             fill = RankGroup)) +
    
    # Use bar geom layer to convey differences between states
    geom_bar(stat = 'identity', 
             width = .6) +
    
    # Color bars based on state rank grouping    
    scale_fill_manual(values = state_colors) +
    
    # Label each state at end of bar whence eye is drawn 
    geom_text(data = function(x) dplyr::filter_(x, ~ RankGroup %in% c('7-10', '11-50')),
              aes(label = State), 
              size = 2, 
              hjust = -.5, 
              vjust = .3) + 
    geom_text(data = function(x) dplyr::filter_(x, ~ RankGroup == 'Top3'), # Magnify size of labels for top 3 states
              aes(label = State), 
              size = 3.7, 
              hjust = -.5, 
              vjust = .3) + 
  
    # Flip axis for portrait orientation
    coord_flip(ylim = c(0, 800), expand = TRUE) +   
    
    # Clearly label and apply muted theme for easy comprehension
    labs(title = 'Companies by state', 
         x = 'State', 
         y = 'Number of companies') +
    theme_minimal() +
    theme(axis.text.y = element_blank(), # Remove axis labels, annotating bar ends instead as eye is drawn there
          axis.ticks = element_blank(),  # Remove chrome / cruft
          legend.position = 'none')  # Remove surplus legend
    
    
```

***

## Question 2

Lets dig in on the state with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries. Create a plot that shows the average and/or median employment by industry for companies in this state (only use cases with full data, use R's `complete.cases()` function.) In addition to this, your graph should show how variable the ranges are, and you should deal with outliers.

```{r fig.with = 9/2, fig.height = 8}

# From the Q1 chart it's evident that NY is the state with the third most companies, following California and Texas.  Accordingly, we chart employees by NY state companies grouped by industry with a horizontal boxplot.  The number of employees per company are plotted on the horizontal axis; industries are plotted on the vertical axis, ranked in descending order by the number of companies in the NY state dataset.  The number of companies are marked along the vertical axis, providing an indication of observations so statistics from low sample industries are intrepreted appropriately.  The distribution density of number of employee per company is depicted by point clouds, with outliers marked in red.  The median and inter-quartile range of employees per company for each industry are represented by blue boxes, and mean employees per company is also plotted in blue to characterize the weight of outliers (for example, business products and services and consumer products and services both have means well above the IQR, reflecting skew of large outliers).  Some outliers have been excluded from view to focus users on drawing clearer contrasts in the distributions between industries.  

# Instantiate function to count number of companies by industry
give.n <- function(x){
  return(c(y = 0, label = length(x)))
}

# Configure data for ggplot
inc[complete.cases(inc),] %>% 
  filter(State == 'NY') %>% 
  select(Name, Industry, Employees) %>% 
  group_by(Industry) %>% 

  # Flag employee counts by company as outliers (> IQR == 'TRUE) or not ('FALSE')
  mutate(Outlier = Employees > median(Employees) + IQR(Employees) * 1.5) %>% 
  
  # Call ggplot
  ggplot(aes(x = reorder(Industry, 
                         X = as.numeric(Employees), 
                         FUN = length),  # Rank industries by number of employees - reference: https://www.rdocumentation.org/packages/DescTools/versions/0.99.19/topics/reorder.factor
             y = Employees)) +
  
    # Use boxplot geom layer to convey median and IQR 
    geom_boxplot(color = 'blue', # Use bold blue to stand out
                 outlier.shape = NA, # Chart outliers as points through separate geom layer
                 coef = 0) +  # Remove whiskers to reduce ink, as median and point clouds better relay distribution - reference: https://stackoverflow.com/questions/18459287/remove-whiskers-in-box-whisker-plot
    
    # Use point geom layers to demonstrate density of distribution
    geom_point(data = function(x) dplyr::filter_(x, ~ Outlier == 'FALSE'), # Geom layer for non-outliers in black 
               size = 2,
               shape = 16, 
               alpha = .2,
               position = position_jitter(w = .3, h = .1)) + # Use alpha and jitter together to convey density and distribution
    geom_point(data = function(x) dplyr::filter_(x, ~ Outlier == 'TRUE'), # Geom layer for outliers in red
               size = 2,
               color = 'red',
               shape = 16, 
               alpha = .15,
               position = position_jitter(w = .15, h = .1)) + # Adjust alpha jitter (less spread for lower density of points)
  
    # Add point at mean to mirror summary statistic cues of boxplot geom layer
    stat_summary(fun.y = 'mean', 
                 geom = 'point', 
                 shape = 21, 
                 size = 3, 
                 color = 'white', # Bold mean's perimeter white for contrast with point clouds 
                 fill = 'blue') + # Use bold blue to mirror IQR boxes

    # Label number of companies for each industry along axis - references: https://stackoverflow.com/questions/28846348/add-number-of-observations-per-group-in-ggplot2-boxplot
    # https://medium.com/@gscheithauer/how-to-add-number-of-observations-to-a-ggplot2-boxplot-b22710f7ef8
    stat_summary(fun.data = give.n, 
                 geom = 'text', 
                 size = 3, 
                 hjust = 1, 
                 position = position_dodge(width = 1)) +
    
    # Flip axis for portrait orientation, adding chart view limits like coord_cartesian layer
    # Avoided scale_y_continuous as it filters out data from stat_summary() - reference: https://github.com/tidyverse/ggplot2/issues/647
    coord_flip(ylim = c(0, 1500), expand = TRUE) +   
    scale_y_continuous(labels = scales::comma,
                       breaks = seq(0, 1600, by = 200)) +
  
    # Clearly label and apply muted theme for easy comprehension
    labs(title = 'NY Companies by Industry',
         subtitle = 'Ranked by Number of Companies',
         x = '', 
         y = '') +
    theme_minimal()


```

***

## Question 3

Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart that makes this information clear. Once again, the distribution per industry should be shown.

```{r fig.with = 9/2, fig.height = 8}

# We use a box-point hybrid plot similar to Q2 in order to chart per capita revenue for all companies grouped by industry.  As this chart is not limited to companies in NY state the dataset is considerably larger than in Q2.  Per capita revenue by company is plotted on the horizontal axis; industries are plotted on the vertical axis and ranked by mean per capita revenue.  In this case, the distribution density of per capita revenue by company is depicted by point clouds, with outliers marked in red.  The median and inter-quartile range of per capita revenue for each industry are represented by blue boxes, and mean per capita revenue is also plotted in blue to characterize the weight of outliers.  As with the Q2 chart, some outliers have been excluded from view to focus users on drawing clearer contrasts in the distributions between industries.

# Instantiate function to count number of companies by industry
give.n <- function(x){
  return(c(y = 0, label = length(x)))
}

# Configure data for ggplot
inc[complete.cases(inc),] %>% 
  select(Industry, Revenue, Employees) %>% 
  mutate(RevPerCap = Revenue / Employees) %>% 
  group_by(Industry) %>% 
  mutate(Outlier = RevPerCap > median(RevPerCap) + IQR(RevPerCap) * 1.5) %>% 
  ungroup() %>% 
  
  # Call ggplot
  ggplot(aes(x = reorder(Industry, 
                         X = as.numeric(RevPerCap), 
                         FUN = mean), # Industries ranked by average per capita revenue
             y = RevPerCap)) + 
    
    # Use boxplot geom layer to convey median and IQR 
    geom_boxplot(color = 'blue', # Use bold blue to stand out
                 outlier.shape = NA, # Chart outliers as points through separate geom layer
                 coef = 0) + # Remove whiskers to reduce ink, as median and point clouds better relay distribution - reference: https://stackoverflow.com/questions/18459287/remove-whiskers-in-box-whisker-plot
    
    # Use point geom layers to demonstrate density of distribution  
    geom_point(data = function(x) dplyr::filter_(x, ~ Outlier == 'FALSE'), # Geom layer for non-outliers in black
               size = 2,
               shape = 16, 
               alpha = .1,
               position = position_jitter(w = .3, h = .1)) + # Use alpha and jitter together to convey density and distribution
    geom_point(data = function(x) dplyr::filter_(x, ~ Outlier == 'TRUE'), # Geom layer for outliers in red 
               size = 2,
               color = 'red',
               shape = 16, 
               alpha = .15,
               position = position_jitter(w = .15, h = .1)) + # Adjust alpha jitter (less spread for lower density of points)
    
    # Add point at mean to mirror summary statistic cues of boxplot geom layer
    stat_summary(fun.y = 'mean', # 
                 geom = 'point', 
                 shape = 21, 
                 size = 3, 
                 color = 'white', # Bold mean's perimeter white for contrast with point clouds 
                 fill = 'blue') + # Use bold blue to mirror IQR boxes

    # Label number of companies for each industry along axis - references: https://stackoverflow.com/questions/28846348/add-number-of-observations-per-group-in-ggplot2-boxplot
# https://medium.com/@gscheithauer/how-to-add-number-of-observations-to-a-ggplot2-boxplot-b22710f7ef80
    stat_summary(fun.data = give.n, 
                 geom = 'text', 
                 size = 3, 
                 hjust = 1, 
                 position = position_dodge(width = 1)) +
    
    # Flip axis for portrait orientation, adding chart view limits like coord_cartesian layer
    # Avoided scale_y_continuous as it filters out data from stat_summary() - reference: https://github.com/tidyverse/ggplot2/issues/647
    coord_flip(ylim = c(0, 2500000), expand = TRUE) + 
    scale_y_continuous(labels = scales::comma) + # Enforce ,000 with numeric axis labels

    # Clearly label and apply muted theme for easy comprehension
    labs(title = 'Per Capita Revenue by Industry',
         subtitle = 'Ranked by per capita revenue',
         y = 'Revenue per Capita',
         x = '') +
    theme_minimal()

```

***

Other references:

- http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization
- https://socviz.co/workgeoms.html